\blue{Numerical Investigation of Multi-Fluid Flows through Dual Saturated-Unsaturated Porous Media - change the title to something close to the main topic of the manuscript ... something with viscous flow instability or viscous fingering)}

### 2017-02-19 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
keep on updating

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Introduction}\label{section:Intro}
\medskip
Numerical investigation and modelling of multiphase flows in porous media have attracted the attention of the scientific community over the past 40 years. Characterization of the flow of fluids through porous media serves as the foundation of reservoir engineering studies. Reservoir engineers and hydro-geologists are dealing with complex geometries, irregular boundaries and heterogeneous geological formations. Oil and gas migration, transport of contaminants through groundwater are some of important physical problem that porous media flows play a crucial role. Underground coal gasification, for the mining industry is an important area of interest, but more recently, also due to the global environmental problems, research focus on the CO$_{\text{2}}$ migration and trapping mechanisms in carbon capture utilization and storage (CCUS) operations~\citep{spycher_2003, chen_2006, self_2012, pruess_1990c, white_1981,jiang_2011}.

\medskip
Out of this mixed and complex various geological formations, multiphase flows through porous media. play the most important role in understanding the afformentioned operations. These often involves the displacement of a non-wetting invading fluid from a porous medium by a wetting fluid. Modeling of multiphase flow, on the other hand is still an enormous technical and computational challenge. In all branches of fluid mechanics, there are the instabilities, also know as flow instabilities. The first distinction should be made between microscopic and macroscopic instabilities. The term microscopic instabilities is used for the phenomena occurring locally; for example, the Helmholtz and Saffman-Taylor instabilities. On the other hand the macroscopic instabilities involve, e.g.the entire two-phase flow system. In this paper, the main focus is kept on macroscopic phenomena. Part of flow instabilities is the viscous flow instabilities, also know as \textit{viscous fingering}. In large scale field operations it is the condition whereby the interface of two fluids, such as oil and water, bypasses sections of reservoir as it moves along, creating an uneven, or fingered, profile. Fingering is a relatively common condition in reservoirs with water-injection wells. The result of fingering is an inefficient sweeping action that can bypass significant volumes of recoverable oil and, in severe cases, an early breakthrough of water into adjacent production wellbores. It can occur both within a single phase or when multiple phases flow through porous media. This instability is essentially governed by the mobility-ratio, MR, which is defined as the ratio of the mobility of the displacing fluid over that of the displaced fluid. In a water-oil system the MR is given by eq.\ref{eq:MR}    However, other important factors also influence the severity of viscous fingering: heterogeneous permeability fields, gravity, anisotropic dispersion, velocity dependence of dispersion, nonmonotonic viscosity profiles, etc. High mobility-ratio displacement processes present very  complex  fingering  patterns.

\KCnote{ \blue{need to move it in a later section}
\begin{equation}
 MR \; = \frac{k_{rw} \; \mu_{w}}{k_{ro} \; \mu_{o}} 
\label{eq:MR}
\end{equation}
}

@@

\medskip
The discretised global mass and force balance equations are solved using a multigrid-like approach ,see \citet{pavlidis2016}.

\begin{eqnarray}
 && \sum_{\alpha=1}^{N_{p}} \int\limits_{\Omega_{CVi}} \; M_{i} \, \frac{\phi\left(S_{\alpha i}^{n+1}\right)-\left(S_{\alpha i}^{n}\right)}{\Delta t} dV  + \nonumber\\
 &&  \oint_{\Gamma_{CVi}} [\theta^{n+1/2}\; n\dot\;v_{p}^{n+1}S_{p}^{n+1}\;+\;(1-)\theta^{n+1/2}\;v_{p}^{n}S_{p}^{n}] d\Gamma \;- \nonumber\\
 &&  \int_{\Omega_{CVi}} M_{i}q_{\alpha}^{n+\theta}dV\; =0
\label{global_mass_balance}
\end{eqnarray}

where $\Omega_{CVi}$ and $\Gamma_{CVi}$ are the volume and boundary of CV i respectively, n is normal vector of the CV and $\theta$ varies smoothly between $0.5$ (corresponding to CN) and $1$ (corresponding to BE) to avoid the introduction of spurious oscillations for large grid Courant numbers. The Eq.\ref{global_mass_balance} can be expressed in amtrix form at time level $n+1$ as

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Brief Overview of Viscous Instabilities}\label{section:ViscousInstabilities}
\medskip
As in many areas of fluid mechanics, the behaviour of the flow for conditions that exceed critical limits, are those that occur under highly critical conditions. Under this assumption, viscous fingering is no exception. It is a viscous driven hydrodynamic instability that results the collapse of the uniform interface between the fluids of different viscosity. This type of instability can be found across many research areas, disciplines and scales. Starting from small scale flow instabilities in Hele-Shaw cells that can provide the basic mathematical model, to coupled chromatographic separation systems for chemical and medical applications and reaching up to large scale reservoir systems,  such as the removal of oil from reservoirs, the potential use of CO$_2$ sequestration and the re-mediation of spills, or the transport of pollutants in aquifers. The primary removal of oil from reservoirs leaves the porous reservoirs with remaining oil that can only be removed by the use of enhanced oil-recovery (EOR). This secondary and tertiary removal of oil from porous beds is required to be as efficient and economically viable as possible.

An accurate finite-element simulator with an adaptive and unstructured 2D grid was used in this study of viscous instabilities. This scheme ha already been describe in detail by \citet{pavlidis2016}. An unfavourable mobility-ratio displacement coupled with the heterogeneity of the porous medium results in viscous instability (fingering),which in turn affects the displacement efficiency of im/miscible processes as well as sweep efficiency and breakthrough time. 
Fingering will occur in miscible displacments when the viscocity ratio is greater than unity. As surface tension becomes weak, the front of the steady finger is susceptible to a viscous-fingering instability by the basic mechanism associated with a less viscous fluid displacing a more viscous one. After a split, each of the new lobes of the finger is stable as a result of their being thinner than the finger from which they split. As a result of \textit{shielding}, one of these lobes will eventually outgrow the others and will then spread to occupy the appropriate width of the cell. In the process, it reaches a width that is again unstable to splitting, and the pattern repeats. Thus, surface tension plays an essential dual role; it must be weak enough for the tip front to be unstable, but it is also the physical force causing the \textit{spreading} and ensuing repeated branching. Important parameter is the mobility (or viscosity) ratio. This is given by Eq.~\ref{eq:mobility-of-fluid} where $K$ is absolute permeability and $k_{r}$ is relative permeability :

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Introduction}\label{section:intro}
\medskip
Numerical investigation and modelling of multiphase flows in porous media have attracted the attention of the scientific community over the past 40 years. Characterization of the flow of fluids through porous media serves as the foundation of reservoir engineering studies. Reservoir engineers and hydro-geologists are dealing with complex geometries, irregular boundaries and heterogeneous geological formations. Oil and gas migration, transport of contaminants through groundwater are some of important physical problem that porous media flows play a crucial role. Underground coal gasification, for the mining industry is an important area of interest, but more recently, also due to the global environmental problems, research focus on the CO$_{\text{2}}$ migration and trapping mechanisms in carbon capture utilization and storage (CCUS) operations~\citep{spycher_2003, chen_2006, self_2012, pruess_1990c, white_1981,jiang_2011}.

\medskip
Out of this mixed and complex various geological formations, multiphase flows through porous media. play the most important role in understanding the afformentioned operations. These often involves the displacement of a non-wetting invading fluid from a porous medium by a wetting fluid. Modeling of multiphase flow, on the other hand is still an enormous technical and computational challenge. In all branches of fluid mechanics, there are the instabilities, also know as flow instabilities. The first distinction should be made between microscopic and macroscopic instabilities. The term microscopic instabilities is used for the phenomena occurring locally; for example, the Helmholtz and Saffman-Taylor instabilities. On the other hand the macroscopic instabilities involve, e.g.the entire two-phase flow system. In this paper, the main focus is kept on macroscopic phenomena. Part of flow instabilities is the viscous flow instabilities, also know as \textit{viscous fingering}. In large scale field operations it is the condition whereby the interface of two fluids, such as oil and water, bypasses sections of reservoir as it moves along, creating an uneven, or fingered, profile. Fingering is a relatively common condition in reservoirs with water-injection wells. The result of fingering is an inefficient sweeping action that can bypass significant volumes of recoverable oil and, in severe cases, an early breakthrough of water into adjacent production wellbores. It can occur both within a single phase or when multiple phases flow through porous media. This instability is essentially governed by the mobility-ratio, MR, which is defined as the ratio of the mobility of the displacing fluid over that of the displaced fluid. %In a water-oil system the MR is given by eq.\ref{eq:MR}.
However, other important factors also influence the severity of viscous fingering: heterogeneous permeability fields, gravity, anisotropic dispersion, velocity dependence of dispersion, nonmonotonic viscosity profiles, etc. High mobility-ratio displacement processes present very  complex  fingering  patterns.

@@

\begin{eqnarray}
 && \sum_{\alpha=1}^{N_{p}} \int\limits_{\Omega_{CVi}} \; M_{i} \; \frac{\phi\left(S_{\alpha i}^{n+1}\right)-\left(S_{\alpha i}^{n}\right)}{\Delta t} dV  + \nonumber\\
 &&  \oint_{\Gamma_{CVi}} [\; \theta^{n+1/2}\; n\dot \; v_{\alpha}^{n+1}S_{\alpha}^{n+1} \; + \; (1-\theta^{n+1/2}) \; v_{\alpha}^{n}S_{\alpha}^{n}] d\Gamma \;- \nonumber\\
 &&  \int_{\Omega_{CVi}} M_{i} \; \mathcal{S}_{cty,\alpha}^{n+\theta} \; dV\; =0
\label{global_mass_balance}
\end{eqnarray}

\noindent where $\Omega_{CVi}$ and $\Gamma_{CVi}$ are the volume and boundary of CV i respectively, \textbf{n} is normal vector of the CV and $\theta$ varies smoothly between $0.5$ (corresponding to Crank-Nicolson method) and $1$ (corresponding to Backward-Euler scheme) to avoid the introduction of spurious oscillations for large grid Courant numbers. In addition to the numerical schemes above the discretised global mass and force balance equations are solved using a multigrid-like approach, see \citet{pavlidis2016}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Brief Overview of Viscous Instabilities}\label{section:ViscousInstabilities}
\medskip
As in many areas of fluid mechanics, the behaviour of the flow for conditions that exceed critical limits may cause instabilities in the system. The flow of a fluid over or next to another heavier fluid, can produce a flow instability. Under this assumption when a fluid is injected in a porous domain saturated with an other fluid trying to displace it, viscous fingering will occur.

\medskip
This viscous driven hydrodynamic instability results the collapse of the uniform interface between the fluids of different viscosity. This type of instability can be found across many research areas, disciplines and scales, from small chemical and medical separation systems to large scale field geological and /or reservoir systems. Starting from the work of \citet{saffman_1959} and \citet{homsy_1987}, and the experimental setup of Hele-Shaw cell by \citet{saffman_1986}, a basic mathematical model has been initially developed to describe the physics of this phenomenon.

\medskip
In viscous fingering, a less viscous fluid (i.e. water) pushes a more viscous fluid (i.e. oil), in a thin rectangular cells (known as Hele-Shaw cells). Under the assumption that fluids remaining completely separated (or immiscible) along an interface, surface tension plays an important role in determining the shape and progress of the fingers. For homogeneous domains fingers will start to develop when surface tension acting on the interface, is opposing towards the change of shape of the interface thus the interface becomes unstable and will collapse, taking a curved shape. In heterogeneous environments this process may also be triggered by permeability differences between smaller regions. The higher the velocity of the injected fluid, the less wide (a tip-splitting behaviour) the finger will be. In addition, pressure differences acting on the interface will produce a net pressure force $\frac{dP}{dx}$ which is a function of the viscosities of the fluids, permeability value and whether gravitational effects are included or not.

As mentioned in section \ref{section:intro}, an unfavourable mobility-ratio (i.e. in a water-oil system, see eq.\ref{eq:MR}, displacement coupled with the heterogeneity of the porous medium results in viscous instability, which in turn affects the displacement efficiency of im/miscible processes.

\begin{equation}
 MR \; = \frac{k_{rw} \; \mu_{w}}{k_{ro} \; \mu_{o}} 
\label{eq:MR}
\end{equation}

Fingering will occur in miscible displacements when the viscocity ratio is greater than unity. %As surface tension becomes weak, the front of the steady finger is susceptible to a viscous-fingering instability by the basic mechanism associated with a less viscous fluid displacing a more viscous one. After a split, each of the new lobes of the finger is stable as a result of their being thinner than the finger from which they split. 
\noindent As a result of \textit{shielding}, one of these fingers will eventually outgrow the others and will then spread to occupy the appropriate width of the cell. In the process, it reaches a width that is again unstable to splitting, and the pattern repeats. Thus, surface tension plays an essential dual role; it must be weak enough for the tip front to be unstable, but it is also the physical force causing the \textit{spreading} and ensuing repeated branching. Important parameter is the mobility (or viscosity) ratio. This is given by Eq.~\ref{eq:mobility-of-fluid} where $K$ is absolute permeability and $k_{r}$ is relative permeability :
